---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  fig.align = 'center'
)
```

# JointAI: Joint Analysis and Imputation of Incomplete Data

[![Travis-CI Build Status](https://travis-ci.org/NErler/JointAI.svg?branch=master)](https://travis-ci.org/NErler/JointAI)
[![CRAN_Status_Badge](https://www.r-pkg.org/badges/version-last-release/JointAI)](https://CRAN.R-project.org/package=JointAI)
[![](https://cranlogs.r-pkg.org/badges/grand-total/JointAI)](https://CRAN.R-project.org/package=JointAI)
[![Download counter](http://cranlogs.r-pkg.org/badges/JointAI)](https://cran.r-project.org/package=JointAI)
[![Rdoc](http://www.rdocumentation.org/badges/version/JointAI)](http://www.rdocumentation.org/packages/JointAI)


The package **JointAI** provides joint analysis and imputation of 
(generalized) linear regression models, (generalized) linear mixed models 
and parametric (Weibull) survival models with incomplete (covariate) data in 
the Bayesian framework.

The package performs some preprocessing of the data and creates a 
[JAGS](http://mcmc-jags.sourceforge.net) model, which will then automatically 
be passed to [JAGS](http://mcmc-jags.sourceforge.net)
with the help of the R package [**rjags**](https://CRAN.R-project.org/package=rjags).

**JointAI** also provides summary and plotting functions for the output.

    
## Installation

You can install **JointAI** from GitHub with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("NErler/JointAI")
```


## Main functions
Currently, there are the following main functions:

``` r
lm_imp()      # linear regression
glm_imp()     # generalized linear regression 
clm_imp()     # cumulative logit model
lme_imp()     # linear mixed model
glme_imp()    # generalized linear mixed model
clmm_imp()    # cumulative logit mixed model
survreg_imp() # parametric (Weibull) survival model
coxph_imp()   # Cox proportional hazards survival model
```

The functions `lm_imp()`, `glm_imp()` and `clm_imp()` use specification similar
to their complete data counterparts `lm()` and `glm()` from base R and `clm()`
from the package [**ordinal**](https://CRAN.R-project.org/package=ordinal).

The functions for mixed models, `lme_imp()`, `glme_imp()` and `clmm_imp()` use
similar specification as `lme()` from the package 
[**nlme**](https://CRAN.R-project.org/package=nlme) (and `clmm2()` from
[**ordinal**](https://CRAN.R-project.org/package=nlme)).

`survreg_imp()` and `coxph_imp()` are missing data versions of `survreg()` and
`coxph()` from the package [**survival**](https://CRAN.R-project.org/package=survival).


Functions `summary()`, `coef()`, `traceplot()` and `densityplot()` provide a summary
of the posterior distribution and its visualization.

`GR_crit()` and `MC_error()` provide the Gelman-Rubin diagnostic for
convergence and the Monte Carlo error of the MCMC sample, respectively.

**JointAI** also provides functions for exploration of the distribution of the
data and missing values, export of imputed values and prediction.


## Minimal Example
### Visualize the observed data and missing data pattern
```{r, fig.width = 8, fig.height = 4, message = FALSE}
library(JointAI)

op <- par(mar = c(2.5, 3, 2.5, 1), mgp = c(2, 0.8, 0))
plot_all(NHANES[c(1, 5:6, 8:12)], fill = '#18bc9c', border = '#2C3E50', ncol = 4, nclass = 30)
par(op)
```


```{r, fig.height = 5}
md_pattern(NHANES, color = c('#2C3E50', '#18bc9c'))
```

### Fit a linear regression model with incomplete covariates

```{r fit_lm1, message = F}
lm1 <- lm_imp(SBP ~ gender + age + WC + alc + educ + bili,
              data = NHANES, n.iter = 500, progress.bar = 'none')
```

### Visualize the MCMC sample
```{r fig.width = 8, fig.height = 4, out.width = '100%', dpi = 300}
traceplot(lm1, col = c('#E74C3C', '#2C3E50', '#18bc9c'), ncol = 4)
```

```{r fig.width = 8, fig.height = 4, out.width = '100%', dpi = 300}
densplot(lm1, col = c('#E74C3C', '#2C3E50', '#18bc9c'), ncol = 4, lwd = 2)
```

### Summarize the Result
```{r}
summary(lm1)
```

```{r}
coef(lm1)

confint(lm1)
```

