---
title: "Parameter Selection"
author: "Nicole Erler"
date: "`r Sys.Date()`"
output: 
  rmarkdown::html_vignette:
    toc: true
vignette: >
  %\VignetteIndexEntry{Monitoring Parameters}
  %\VignetteEncoding{UTF-8}
  %\VignetteEngine{knitr::rmarkdown}
editor_options: 
  chunk_output_type: inline
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.align = 'center'
)
library(JointAI)
```

## Monitoring parameters
**JointAI** uses JAGS for performing the MCMC sampling.
Since JAGS only saves the values of MC chains for those parameters/variables for
which the user has specified that they should be monitored, this is also the
case in **JointAI**.

For this purpose, `lm_imp()`, `glm_imp()` and `lme_imp()` have an argument
`monitor_params`.

`monitor_params` takes a named list (often a named vector also works) with
the following possible entries:

```{r, echo = F}
tab <- rbind(
c("`analysis_main`", "`betas`, `tau_y` and `sigma_y`"),
c("`analysis_random`", "`ranef`, `D`, `invD`, `RinvD`"),
c('`imp_pars`', '`alphas`, `tau_imp`, `gamma_imp`, `delta_imp`'),
c('`imps`', 'imputed values'),
c('`betas`', 'regression coefficients of the analysis model'),
c('`tau_y`', 'precision of the residuals from the analysis model'),
c('`sigma_y`', 'standard deviation of the residuals from the analysis model'),
c('`ranef`', 'random effects'),
c('`D`', 'covariance matrix of the random effects'),
c('`invD`', 'inverse of `D`'),
c('`RinvD`', 'scale matrix in Wishart prior for `invD`'),
c('`alphas`', 'regression coefficients in the imputation models'),
c('`tau_imp`', 'precision parameters of the residuals from imputation models'),
c('`gamma_imp`', 'intercepts in ordinal imputation models'),
c('`delta_imp`', 'increments of ordinal intercepts'),
c('`other`', 'additional parameters')
)
colnames(tab) = c('name/key word', 'what is monitored')

knitr::kable(tab)
```

Each of the key words works as a switch.

In the following, we use the [NHANES](https://nerler.github.io/JointAI/reference/NHANES.html)
data (for examples in cross-sectional data) and the 
dataset [simLong](https://nerler.github.io/JointAI/reference/simLong.html) for
examples in longitudinal data.

In many of the examples we use `n.adapt = 0`
(and `n.iter = 0`, which is the default) in order to prevent the MCMC sampling 
to save some time.
`mess = FALSE` is used to suppress messages that are not of interest in this
vignette. `progress.bar = 'none'` prevents printing of the progress of the MCMC
sampling, which results in lengthy output in the vignette, but is useful in 
practice.

### Parameters of the analysis model
The default setting is `monitor_params = c(analysis_main = TRUE)`, i.e., 
only the main parameters of the analysis model are monitored and 
monitoring is switched off for all other parameters.

The main parameters are the regression coefficients of the analysis model
(`beta`) and the residual standard deviation (`sigma_y`,  and precision `tau_y`).

The function `parameters()` returns the parameters that are specified to be followed
(even for models where no MCMC sampling was performed, i.e. when `n.iter = 0` and
`n.adapt = 0`).

For example: 
```{r lm1_1}
lm1 <- lm_imp(SBP ~ gender + WC + alc + creat,
              data = NHANES, n.adpat = 0, mess = FALSE)
parameters(lm1, mess = FALSE)
```



### Imputed values & parameters of the imputation models
To generate (multiple) imputed datasets that can be used for further analyses,
the imputed values need to be monitored. This can be done by setting
`monitor_params = c(imps = TRUE)`.

```{r lm2_1}
lm2 <- lm_imp(SBP ~ age + WC + alc + smoke + occup,
              data = NHANES, n.adapt = 0, mess = FALSE,
              monitor_params = c(imps = TRUE, analysis_main = FALSE),
)
parameters(lm2, mess = FALSE)
```

**JointAI** uses a number of different design matrices to store different
types of variables. The matrix `Xc` is the design matrix of cross-sectional
covariates. For categorical incomplete variables (with more than 2 categories)
the original variable is stored in the matrix `Xcat` and `Xc` contains the
corresponding dummy coded variables.
Hence, the imputed values of continuous and binary variables are elements of
`Xc` and imputed values of categorical variables are elements of `Xcat`.


The parameters of the models for the incomplete variables can be selected with
`monitor_params = c(imp_pars = TRUE)`. This will set monitors for the
regression coefficients (`alpha`) and other parameters, such as precision
(`tau_*`) and intercepts & increments (`gamma_*` and `delta_*`) in cumulative
logit models.

```{r lm3_1}
lm3 <- lm_imp(SBP ~ age + WC + alc + smoke + occup,
              data = NHANES, n.adapt = 0, mess = FALSE,
              monitor_params = c(imp_pars = TRUE, analysis_main = FALSE),
)
parameters(lm3, mess = FALSE)
```

#### Side note: Getting information about of the imputation models
An overview of the imputation models used, including the names of the 
parameters and the hyperparameters can be obtained with
```{r list_lm2}
list_impmodels(lm2)
```

#### Side note: How to extract imputed datasets
Imputed datasets can be extracted and exported with the function `get_MIdat()`.
A completed dataset is created by taking the imputed values from a randomly 
chosen iteration of the MCMC sample (transforming them back to the original scale,
if scaling had been performed during the MCMC sampling)
and filling them into the original, incomplete data.

`get_MIdat()` returns a long-format `data.frame` containing the imputed datasets
(and possibly the original data) stacked onto each other. The imputation number
is given in the variable `Imputation_`, column `.id` contains a newly created
id variable for each observation in cross-sectional data (multi-level data should
already contain an id variable).

`get_MIdat()` takes the arguments:
```{r, echo = FALSE}
tab <- rbind(object = 'a JointAI object',
             m = "number of datasets to be created",
             include = "logical; should the original data be included?",
             start = paste0("the first iteration that may be randomly chosen",
                            " (i.e., all previous iterations are discarded as burn-in)"),
             seed = "optional seed value in order to make the random selection
                     of iterations reproducible",
             export_to_SPSS = paste0("logical; should the datasets be exported", 
                                     " to SPSS, i.e., written as .txt and .sps file? ",
                                     "If `export_to_SPSS = FALSE` (default) the",
                                     " imputed data is only returned `data.frame`"),
             filename = 'the name of the .txt and .sps files',
             resdir = 'directory the files are exported to') 

tab <- cbind(paste0("`", rownames(tab), "`"), tab)
colnames(tab) <- c('argument', 'explanation')

knitr::kable(tab, row.names = FALSE)
```

### Random effects
For mixed models, `analysis_main` also includes the random effects covariance
matrix `D`:
```{r}
lme1 <- lme_imp(bmi ~ age + EDUC, random = ~age|ID, 
                data = simLong, n.adapt = 0, mess = FALSE)
parameters(lme1, mess = FALSE)
```

Setting `analysis_random = TRUE` will switch on monitoring
for the random effects (`ranef`), random effects covariance matrix (`D`),
inverse of the random effects covariance matrix (`invD`) and the diagonal of the
scale matrix of the Wishart-prior of `invD` (`RinvD`). 

```{r}
lme2 <- lme_imp(bmi ~ age + EDUC, random = ~age|ID,
                data = simLong, n.adapt = 0, mess = FALSE,
                monitor_params = c(analysis_random = TRUE))
parameters(lme2, mess = FALSE)
```


It is possible to select only a subset of the random effects parameters by 
specifying them directly, e.g.
```{r}
lme3a <- lme_imp(bmi ~ age + EDUC, random = ~age|ID,
                data = simLong, n.adapt = 0, mess = FALSE,
                monitor_params = c(analysis_main = TRUE, RinvD = TRUE))
parameters(lme3a, mess = FALSE)
```

or by switching unwanted parts of `analysis_random` off, e.g.

```{r}
lme3b <- lme_imp(bmi ~ age + EDUC, random = ~age|ID,
                data = simLong, n.adapt = 0, mess = FALSE,
                monitor_params = c(analysis_main = TRUE,
                                   analysis_random = TRUE,
                                   RinvD = FALSE,
                                   ranef = FALSE))
parameters(lme3b, mess = FALSE)
```


### Other parameters
The element `other` in `monitor_params` allows to specify one or multiple parameters
additional parameters to be monitored. When `other` is used with more than one
element, `monitor_params` has to be a list.

Here, we monitor the probability to be in the `alc>=1` group for subjects 1
through 3 and the expected value of the distribution of `creat` for the first
subject.
```{r}
lm4 <- lm_imp(SBP ~ gender + WC + alc + creat,
              data = NHANES, n.adapt = 0, mess = FALSE, progress.bar = 'none',
              monitor_params = list(analysis_main = FALSE,
                                    other = c('p_alc[1:3]', "mu_creat[1]")))
parameters(lm4)
```


## Subsets of Parameters for Plots, Summaries, etc.
The functions `summary()`, `traceplot()`, `densplot()`, `GR_crit()` and `MC_error()`
all have an argument `subset`. This argument allows to select a subset of 
parameters to be shown in the output.
Especially when not only the parameters of the main
analysis model are followed, but also, for example, imputed values, looking at
a subset may be desirable.

`subset` follows the same logic as `monitor_params` described above.

By default, only the parameters of the main analysis model are displayed if they
were monitored:
```{r, fig.width = 7, fig.height = 6, cache = TRUE}
# Run a model monitoring analysis parameters and imputation parameters
lm5 <- lm_imp(SBP ~ gender + WC + alc + creat,
              data = NHANES, n.iter = 100, mess = FALSE, progress.bar = 'none',
              monitor_params = c(imp_pars = TRUE))

# model summary
summary(lm5)

# traceplot of the MCMC sample
traceplot(lm5)

# density plot of the MCMC sample
densplot(lm5)

# Gelman-Rubin criterion
GR_crit(lm5)

# Monte Carlo Error of the MCMC sample
MC_error(lm5)
```

When `analysis_main` was not switched on, all parameters are displayed by default:
```{r, cache = TRUE, fig.height = 1.5}
# Re-run the model from above, now creating MCMC samples
lm4 <- lm_imp(SBP ~ gender + WC + alc + creat,
              data = NHANES, n.iter = 100, mess = FALSE,  progress.bar = 'none',
              monitor_params = list(analysis_main = FALSE,
                                    other = c('p_alc[1:3]', "mu_creat[1]")))

traceplot(lm4, ncol = 4)
```

### Select a subset of the variables to display
To display other parts of the MCMC sample, `subset` needs to be specified:
```{r GRcrit_lm5}
# we use lm5 from above
GR_crit(lm5, subset = c(analysis_main = FALSE, imp_pars = TRUE))
```

To select only some of the parameters, they can be specified directly by
name via the `other` element of `subset`:
```{r trace_lm5, fig.width = 5, fig.height = 2, out.width = "60%"}
traceplot(lm5, subset = list(other = c('creat', 'alc>=1')), ncol = 2)
```


To display a subset of the imputed values, the `other` can be specified.
```{r lm2_2, cache = TRUE, fig.height = 2, fig.width = 5, out.width = "60%"}
# Re-run the model from above, now creating MCMC samples
lm2 <- lm_imp(SBP ~ age + WC + alc + smoke + occup,
              data = NHANES, n.iter = 100, mess = FALSE,  progress.bar = 'none',
              monitor_params = c(imps = TRUE, analysis_main = FALSE),
)

# select only imputed values for 'WC' (3rd column of Xc)
sub3 <- grep('Xc\\[[[:digit:]]+,3\\]', parameters(lm2), value = TRUE)
sub3

traceplot(lm2, subset = list(other = sub3), ncol = 2)
```


### Random subset of subject-specific values
When the number of imputed values is larger, or in order to check convergence
of random effects, it may not be feasible to plot all traceplots. 
In that case, a random subset of, for instance the random effects, can be selected:
```{r, fig.height = 4}
lme4 <- lme_imp(bmi ~ age + EDUC, random = ~age|ID,
                data = simLong, n.iter = 100, mess = FALSE, progress.bar = 'none',
                monitor_params = c(analysis_main = FALSE, ranef = TRUE))

# exract random intercepts
ri <- grep('^b\\[[[:digit:]]+,1\\]$', colnames(lme4$MCMC[[1]]), value = T)

# extract random slopes
rs <- grep('^b\\[[[:digit:]]+,2\\]$', colnames(lme4$MCMC[[1]]), value = T)

# plot the chains of 8 randomly selected random intercepts
traceplot(lme4, subset = list(other = sample(ri, size = 8)), ncol = 4)

# plot the chains of 8 randomly selected random slopes
traceplot(lme4, subset = list(other = sample(rs, size = 8)), ncol = 4)
```


